---
title: Population Trend Analysis
author: 'Claire Madden'
date: '2020-01-09'
slug: hares
categories: []
tags:
  - data
subtitle: 'Evaluating Trends in Snowshoe Hare Populations in the Bonanza Creek Experimental Forest, Alaska'
summary: 'Population Trend Analysis of Snowshoe Hares in Alaska'
authors: []
lastmod: '2020-01-09T15:02:21-08:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)

# load packages

library(tidyverse)
library(kableExtra)
library(janitor)
library(lubridate)
library(tidyr)
library(ggfortify)
library(RColorBrewer)
```
Image of a sneaky snowshoe hare in Alaska. Image source: National Parks Service (https://www.nps.gov/gaar/learn/nature/snowshoe-hare.htm).

<br> 

Snowshoe hares have been identified as a 'keystone' prey species in northern boreal forests. This study, conducted in the Bonanza Creek Experimental Forest just east of Fairbanks, Alaska, tracked snowshoe hare populations to generate quantitative data that could be used to better explain and predict vegetation and preditor responses to fluctuations in hare populations. Snowshoe hares are known to experience population fluctuations every 8-11 years. This trend is demonstrated by fluctuations in observation counts during the monitoring period; counts of hare observations declined between 1999 and 2002, and then showed a steady increase from ~2002 to ~2010 (Figure 1, Table 1) (Kielland, Chapin & Ruess, 2017). 

<br>

```{r}

# read in the data 
hares<- read_csv("showshoe_lter.csv") 

hares_mf <- hares %>% 
  mutate(sex = str_to_lower(sex)) %>% # change all entries in sex column to lower case
  filter(sex == "m" | sex == "f") %>% # remove all observations with NA or ? in sex
  mutate(date = mdy(date)) %>% # change date format
  mutate(full_date = date) %>% # retain date column
  separate(date, into = c("year", "month", "day")) %>% #create new columns with parsed date info
  mutate(month = as.numeric(month)) %>% 
  mutate(season = ifelse(month %in% c(1:4, 11, 12), "winter", "summer")) %>% # organize months into "summer" and "winter"
  mutate(year = as.numeric(year))
  
# seems like there are a lot more observations in the summer than in the winter so maybe not an interesting thing to use for visualization

# explore graph
ggplot(data = hares_mf, aes(x = weight, y = hindft))+
  geom_point()+
  facet_wrap(~sex)

# this is not that interesting to me

```

```{r}

# try something else that may be more interesting

hares_hist <- ggplot(data = hares_mf, aes(x = year))+ # plot observation frequency by sex
  geom_bar(aes(fill = sex), position = "dodge")+
  scale_fill_manual(values = c("peru", "peachpuff2"), labels = c("Female", "Male"))+ # set colors and legend names
  scale_x_continuous(expand = c(0,0),
                     breaks = seq(1998, 2012, 1))+ # scale x axis
  theme_minimal()+
  labs(x = "Year of Observation", # set axes labels
       y = "Observation Count")+
  theme(legend.title = element_blank()) # remove legend title

hares_hist


```
        
**Figure 1. Snowshoe Hare Monitoring Observations in the Bonanza Creek Experimental Forest Between 1998 - 2012.**        
Total number of observations of female and male snowshoe hares are recorded each year during the monitoring period. Snowshoe hare population fluctuations within the experimental area are demonstrated by the frequency of monitoring observations throughout the 15 year study. Data source: Kielland, Chapin & Ruess, 2017.

<br>

Early in the study period (1998 - 2002) observations of males frequently exceeded observations of females, while later in the study period (2003 - 2012) females were observed more frequently than males in every year. Additionally, overall observations declined early in the study period, increased after 2002, and began to decline again in 2010. This may be demonstrative of a larger known pattern of snowshow hare population fluctuations occuring every 8-11 years (Kielland, Chapin & Ruess, 2017). 

<br> 
```{r}

# make a summary table for females
hare_summary_f <- hares_mf %>% 
  filter(sex == "f") %>% 
  drop_na(weight, hindft) %>% # take out an values to be able to calculate mean
  group_by(year) %>% # organize by year and sex
  summarize(
    count = length(sex), # number of observations
    avg_weight = round(mean(weight), digits = 0))# mean weight
    
# make a summary table for males
hare_summary_m <- hares_mf %>% 
  filter(sex == "m") %>% 
  drop_na(weight, hindft) %>% # take out an values to be able to calculate mean
  group_by(year) %>% # organize by year and sex
  summarize(
    count = length(sex), # number of observations
    avg_weight = round(mean(weight), digits = 0)) # mean weight

# combine the two tables!
hare_summary <- merge(hare_summary_f, hare_summary_m, by = "year")

# format into a single pretty table
hare_table <- kable(hare_summary, align = "c", col.names = c("Year", "Count", "Average Weight (g)", "Count", "Average Weight (g)"), caption = "Table 1. Sumamry Statistics for Snowshoe Hare Monitoring Observations at Bonanza Creek Experimental Forest Between 1998 - 2012. Annual observation counts and average weight of female and male snowshoe hares are reported for each of the years during the monitoring period.") %>% 
  add_header_above(c(" " = 1, "Female" = 2, "Male" = 2)) %>% 
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)

hare_table

```

<br> 

Reference:             

Kielland K., F. S. Chapin, R. W. Ruess. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14. 






