---
title: Timeseries Analysis
author: 'Claire Madden'
date: '2020-01-09'
slug: timeseries
categories: []
tags:
  - data
subtitle: 'Timeseries Analysis of Steelhead Trout Passage at Bonneville Dam on the Columbia River, Oregon from 1939 - 2019'
summary: 'Timeseries Analysis of Steelhead Trout Passage at Bonneville Dam'
authors: []
lastmod: '2020-01-09T15:02:21-08:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.width = 8.5)
```

### Steelhead Trout at Bonneville Dam

Steelhead trout in the Columbia River Basin are ESA listed threatened species. These fish are especially vulnerable given their migratory behavior. Steelhead spend the majority of their lives in the ocean, but travel upstream in rivers across the west coast to freshwater spawning grounds once a year. Human alterations to natural river systems have impacted the fish population's ability to maintain this migratory behavior, and dams in particular create a significant barrier to fish passage (NOAA Fisheries). The Bonneville Dam, on the Columbia River in Oregon, has been outfitted with a fish ladder in order to facilitate annual fish passage to freshwater spawning grounds. The following is an exploration of fish passage counts at Bonneville Dam between 1939 and 2019. 



```{r}

# load packages 
library(tidyverse)
library(tsibble)
library(fable)
library(fabletools)
library(feasts)
library(forecast)
library(janitor)
library(lubridate)
library(paletteer)

# read in data
fish_data <- read_csv("cbr_fish_passage_bonneville_allyrs_steelhead.csv") %>% 
  clean_names()

# data explorations

#View(fish_data)
#unique(fish_data$parameter) # only Stlhd contained in the dataset
#unique(fish_data$datatype) # only Adult Passage recorded
#class(fish_data$mm_dd) #character format 

```


```{r}
# create a new dataframe with additional date formats
fish_clean <- fish_data %>% 
  mutate(date = paste(mm_dd, year, sep = "-")) %>% # combine date info into a single column
  mutate(ts_date = lubridate::parse_date_time(date, "dmy")) %>% # turn into a recognizable format 
  mutate(ts_date = as.Date(ts_date)) %>% # change class to date
  mutate(month_sep = yearmonth(ts_date)) %>% #not sure if i need this step but leaving it in case
  drop_na(month_sep) %>% # remove na values from month_sep (all are on February 29th which must be considered a 'nonsensical' date in tsibbles brain..?)
  drop_na(value) # drop na observations in value column

# create another dataframe with parsed date data for use later
fish_parse <- fish_clean %>% 
  mutate(new_date = ts_date) %>% # duplicate date column
  separate(new_date, into = c("year", "month", "day"), sep = "-") %>% #seperate date components
  mutate(year = as.numeric(year), # make all date components recognized as numeric (not character)
         month = as.numeric(month),
         day = as.numeric(day))
  
```



```{r}
# timeseries plot of original observations (daily data)
fish_ts_daily <- ggplot(data = fish_clean, aes(x = ts_date, y = value))+
  geom_line()+
  theme_bw()+
  labs(x = "Date",
       y = "Count of Adult Steelhead Trout \nPassing Bonneville Dam")+
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,35000),
                     labels = scales::comma)

fish_ts_daily # pretty tough to see any details on this


```
                               
**Figure 1. Daily Steelhead Trout Passage Counts at Bonneville Dam (1939 - 2019).** Timeseries plot showing daily counts of steelhead trout passing Bonneville Dam between 1939 and 2019. The maximum single day count occured on August 13th 2009 with 34,053 fish recorded. 


```{r}
# check out observations since 2000 to see more detail
fish_since2000 <- fish_clean %>% 
  filter(year>=2000) # remove observations made before 2000

# graph that!
fish_ts_2000 <- ggplot(data = fish_since2000, aes(x = ts_date, y = value))+
  geom_line()+
  theme_bw()+
  labs(x = "Date",
       y = "Count of Adult Steelhead Trout \nPassing Bonneville Dam")+
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,35000),
                     labels = scales::comma)

fish_ts_2000 # this one is a lot better for seeing seasonality/yearly peaks
```
                                
**Figure 2. Daily Steelhead Trout Passage Counts at Bonneville Dam (2000 - 2019).** Timeseries plot showing daily counts of steelhead trout passing Bonneville Dam between 2000 and 2019. The annual peaks shown in the timeseries correspond with the spawning season when thousands of steelhead travel up the Columbia River to freshwater spawning grounds. 


```{r}

# need to make a monthly summary of values to make a season plot

fish_month <- fish_clean %>% 
  group_by(month_sep) %>% 
  summarize(month_agg = sum(value)) %>% 
  mutate(year_month = tsibble::yearmonth(month_sep),
         month = month(year_month, label = TRUE),
         year = year(year_month))

fish_month_ts <- as_tsibble(fish_month, index = year_month) %>% # coerce to tsibble to check gg_subseries etc
  tsibble::fill_gaps() # this step is required in order to run gg_subseries

fish_month_ts %>% autoplot(month_agg) # hard to tell this from the daily timeseries
fish_month_ts %>% gg_subseries(month_agg) # subseries looks as i would expect!
#fish_month_ts %>% gg_season(month_agg) # gg_season still not working

# plot monthly total timeseries
month_season_plot <- ggplot(data = fish_month, aes(x = month, y = month_agg, group = year)) +
  geom_line(aes(color = year))+
  scale_color_paletteer_c("ggthemes::Orange-Blue Diverging")+
  theme_bw()+
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,400000),
                     labels = scales::comma)+
  labs(x = "Month",
       y = "Monthly Total Steelhead Passage \n at Bonneville Dam",
       color = "Observation Year")
  
month_season_plot


```
                             
**Figure 3. Monthly Total Steelhead Trout Passage Counts at Bonneville Dam (1939 - 2019).** This season plot displays the total monthly counts of steelhead trout passing Bonneville Dam for every year between 1939 and 2019. Each line represents a single year contained within the dataset, with warm colors representing earlier years in the period of record and cool colors representing more recent years in the period of record. The color gradient displays an increasing trend in monthly total steelhead counts in recent years (blue lines). 


```{r}
# annual passages

fish_year <- fish_clean %>% 
  group_by(year) %>% # organize by year
  summarize(annual_agg = sum(value)) %>% # make annual totals column
  mutate(year = year) # keep a column with year

# graph it!
fish_year_ts <- ggplot(data = fish_year, aes(x = year, y = annual_agg)) +
  geom_line()+
  scale_y_continuous(expand = c(0,0),
                     limits = c(0, 650000),
                     breaks = seq(0, 650000, 100000),
                     labels = scales::comma)+
  scale_x_continuous(expand = c(0,0),
                     breaks = seq(1940, 2025, 10))+
  labs(x = "Year",
       y = "Count of Total Annual Steelhead Trout\n Passing Bonneville Dam")+
  theme_bw()



fish_year_ts

```
                               
**Figure 4. Annual Steelhead Trout Passage Counts at Bonneville Dam (1939 - 2019).** Timeseries plot showing total annual counts of steelhead trout passing Bonneville Dam between 1939 and 2019. This timeseries confirms the trend in Figure 3, indicating that total counts of steelhead passing the dam have an increasing trend in recent years. 

<br>
<br>

References:

NOAA Fisheries. (n.d.) Species Directory: Steelhead Trout. Accessed 2/19/20 from: https://www.fisheries.noaa.gov/species/steelhead-trout#overview

Columbia River DART, Columbia Basin Research, University of Washington. (2019). Adult Passage Graphics & Text. Available from http://www.cbr.washington.edu/dart/query/adult_graph_text


